---
title: "Microbialdiversity1.0: Diversity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Microbialdiversity1.0: Diversity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
Introduction
This vignette demonstrates how to use the Microbialdiversity1.0 package to calculate and compare various diversity indices commonly used in ecological and microbiome research. The package includes:
Shannon diversity index
Simpson diversity index
Inverse Simpson diversity index
Observed OTUs (Operational Taxonomic Units)

The Microbialdiversity1.0 package provides a user-friendly way to perform these calculations.

#Setup
```{r setup}
library(Microbialdiversity1.0)
```
Example Dataset
The package includes an example dataset called microbiome_data. It contains:

abundances: A matrix of microbial counts (rows: species, columns: samples).
metadata: Metadata for samples (e.g., group assignments).
Load the dataset:
```{r}
data(microbiome_data)

# Inspect the structure of the dataset
str(microbiome_data)
```
Diversity Indices Calculation
The package provides functions to calculate different diversity indices for a single sample.

Shannon Diversity
The Shannon diversity index measures species richness and evenness:
```{r}
# Calculate Shannon diversity for the first sample
sample1 <- microbiome_data$abundances[, 1]
shannon_index <- calculate_shannon(sample1)
shannon_index

```
Simpson Diversity
The Simpson index measures the probability that two randomly selected individuals belong to the same species:
```{r}
# Calculate Simpson diversity
simpson_index <- calculate_simpson(sample1)
simpson_index
```

Inverse Simpson Diversity
The Inverse Simpson index provides an inverse measure of dominance:
```{r}
# Calculate Inverse Simpson diversity
inv_simpson <- calculate_inverse_simpson(sample1)
inv_simpson
```

Observed OTUs
The Observed OTUs metric counts the number of unique species:
```{r}
# Calculate Observed OTUs
otus <- calculate_observed_otus(sample1)
otus
```
Comparing Diversity Between Groups
The package includes functions to compare diversity indices between groups.

Compare Shannon Diversity
```{r}
# Subset data for Control and Treatment groups
control_samples <- microbiome_data$abundances[, microbiome_data$metadata$condition == "Control"]
treatment_samples <- microbiome_data$abundances[, microbiome_data$metadata$condition == "Treatment"]

# Compare Shannon diversity between groups
shannon_comparison <- compare_shannon_two(control_samples, treatment_samples)
shannon_comparison
```

Compare Simpson Diversity
```{r}
# For comparing two groups
simpson_comparison <- compare_simpson_two(control_samples, treatment_samples)

# For multiple groups, just pass the abundance matrix
simpson_comparison <- compare_simpson_multiple(microbiome_data$abundances)
```
Interpretation Guide
Shannon Diversity Index:Ranges from 0 to ln(S), where S is the number of species. Higher values indicate greater diversity. It also Considers both richness and evenness. Its also  Sensitive to rare species.
Simpson Diversity Index (1-D):  Ranges from 0 (low diversity) to 1 (high diversity). This inddex emphasizes dominant species and is less sensitive to rare species. The simpson Value represents probability that two randomly selected individuals belong to different species.
Inverse Simpson Index:Ranges from 1 to the number of species. Higher values indicate greater diversity. It gives more weight to abundant species, and its less influenced by rare species.
Observed OTUs: Its a simple count of species present (abundance > 0). It measures species richness. It does not consider relative abundances. Its useful for comparing basic community composition

P-values
P < 0.05 indicates significant differences between communities

Multiple comparison matrices show pairwise p-values
Diagonal values are always 1 (comparing community with itself)

Conclusion
The Microbialdiversity1.0 package provides a comprehensive toolkit for analyzing community diversity using multiple indices. Each metric offers different insights into community structure:Shannon index for overall diversity, Simpson index for dominant species, Inverse Simpson for abundance-weighted diversity, Observed OTUs for species richness.
Choose the appropriate metric based on your research questions and the aspects of community diversity you want to emphasize
